---
title: "DATRAS Webservice and Field-List Gaps"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Why this exists

The [other article](recent-fixes.html) in this pair covers code bugs -- places
where, given correct input, this package's own logic produced a wrong result. This
one is about something upstream of that entirely: real gaps in what the DATRAS
webservice itself provides, and in the field list it publishes to describe its own
data. Neither is fixable by changing this package's logic; both are patched here as
interim, escapable workarounds -- see the "webserver/schema gap" category in
`AGENTS.md` for the general framework this falls under.

This is written up separately because the two problems are genuinely different in
kind: one is missing behaviour on ICES's server, the other is missing or wrong
metadata *about* behaviour that already works fine.

## The webservice itself is incomplete

The DATRAS webservice (`https://datras.ices.dk/WebServices/DATRASWebService.asmx`)
exposes 25 operations in total (checked live, 2026-07-22). ICES is evidently
mid-migration on header naming for at least one record type: `getHLdata` has a
sibling, `getHLdataNewHeaders`, returning the same data under updated column names.
But that's the *only* such pair -- there is no `getHHdataNewHeaders`,
`getCAdataNewHeaders`, `getDatrasFieldListNewHeaders`, or equivalent for FL, LT,
Indices, or either CPUE endpoint. The migration, in other words, is real but only
about 4% rolled out across the service's surface.

icesDatras itself doesn't call the one endpoint that does exist in its new form --
every function in this package still uses the old-style operations, translating
names client-side via `getDatrasFieldList()` instead. Cross-checking
`getHLdataNewHeaders` directly against that client-side translation (real NS-IBTS
2022 Q1 data) found 28 of 29 columns agreeing exactly -- good validation that the
translation approach is generally sound -- and one real mismatch, detailed below.

Separately, `getCatchWgt()` has no true server-side equivalent at all: there is no
"get catch weight" operation on the service. The function fabricates one client-side,
fetching HH and HL and summing `CatCatchWgt` per haul in R. Anyone re-implementing
DATRAS access from scratch should know this going in -- it looks like a single
endpoint from the outside but isn't one.

## The field list is incomplete or ambiguous

`getDatrasFieldList()` is the service's own description of its data: which fields
exist per record type, their old and new names, and their intended type. In
practice it has several distinct kinds of gaps against what the service actually
returns:

1. **Plain formatting inconsistencies** -- `DataFormat` uses `"float"` in some rows
   and `"decimal"` in others for what should be the same type; `Year` isn't
   consistently marked `"int"`; a couple of FA fields have the wrong `DataFormat`
   or a missing `Description` entirely.
2. **Wrong `FieldNameOld` values** -- for three fields LT shares with HH
   (`Platform`, `StationName`, `HaulNumber`), the list's `FieldNameOld` column simply
   echoes the *new* name instead of the actual old column name the service returns
   with `new_names = FALSE`.
3. **Missing rows entirely, at record-type scale** -- the list covers only LT's
   ~22 upload-spec fields, but `getLTassessment()` actually returns on the order of
   40 additional HH-style columns (coordinates, gear metrics, haul metadata) with no
   corresponding entry at all.
4. **A whole record type absent** -- FL isn't in the list at all, despite
   `getFlexFile()` returning every HH column plus several FL-specific swept-area
   ones.
5. **Server-added columns with no spec entry** -- the live database appends
   `DateofCalculation` (HH/HL/CA) and a validated Aphia code column to HL/CA
   responses, neither of which the upload-spec field list ever mentions.
6. **A field matching neither documented name** -- CA's age column comes back as
   `Age`, matching neither the list's new name (`IndividualAge`) nor its documented
   old name (`AgeRings`) -- confirmed live, even with `new_names = FALSE`.
7. **A locally-invented name diverging from ICES's own real direction** -- this is
   not something `ices-tools-prod/icesDatras` (upstream, the "mother repo") ever
   did: the official field list has no entry for this column at all. This *fork's*
   own local hot-fix block had filled that gap by translating `Valid_Aphia` to a
   made-up canonical name, `aphia`. The one place ICES's real header migration
   touches this field (`getHLdataNewHeaders`) still calls it `Valid_Aphia`,
   unchanged -- so the invented name diverged from ICES's own actual direction, not
   from anything upstream chose.

Items 1-5 predate this session and originate from this fork's own hot-fix block, not
from upstream; items 6-7 were found and patched here.

## Discrepancies across tables, not just within one

Everything above is one record type's own field list being wrong or incomplete. A
related, likely larger issue sits one level up: ICES's own data model has
inconsistencies *across* HH/HL/CA/FL/LT/Indices/CPUE* -- in both naming conventions
and typing -- that a client library has no business trying to resolve on its own.

A broader, separate investigation comparing several independent DATRAS-consuming
tools directly landed on a useful distinction: HH/HL/CA are institute
*measurements* -- interpreting them (units,
codes, key uniqueness, naming) has exactly one right answer, so every downstream
tool solving that independently is waste, not legitimate scientific difference.
Derived products (FL, indices, spread models) are a different case and can
legitimately diverge -- that's normal practice, not a bug.

Two concrete cases from that investigation bear directly on this package:

- **A value's actual meaning sometimes depends on a different column, from a
  different table.** `NumberAtLength`/`SpeciesCategoryWeight` in HL only mean
  something once joined against `DataType`/`HaulDuration` from HH -- exactly the
  correction this package's own `getCatchWgt()` gets wrong for `DataType == "C"`
  (see the companion article). `LengthClass` similarly means nothing without its
  companion `LngtCode`.
- **Old-name/new-name inconsistency** was named, independently of anything in this
  package, as one of a handful of things every serious downstream DATRAS tool ends
  up re-solving on its own. This package's own field-list hot-fix (documented
  above) is one instance of that pattern, not an isolated case.

None of this is fixable by patching `icesDatras`. A client-side patch can only ever
correct the symptom for this package's own users -- the actual fix is ICES settling
on one answer at the data-model level, which is a question for a Working Group
(IBTSWG has been suggested as the right venue, since DATRAS already briefed them on
related work), not something a webservice client can resolve for itself.

## icesVocab: a related, currently-unused source of field/vocabulary truth

`icesVocab` -- a separate ICES package, not currently consulted by `icesDatras` for
anything -- governs the *values* many DATRAS fields are allowed to take (gear
codes, data-type codes, validity flags, and more) via `getCodeList(code_type, ...)`.
Two things about it are relevant here, both confirmed directly rather than assumed:

- It's piecemeal in the same way `getDatrasFieldList()` is: `getCodeList()` takes
  exactly one `code_type` per call with no bulk/"all" option, and
  `getCodeTypeList()` returns type *names* only, not their code values. Even ICES's
  own official DATSU submission-validation code (`icesDatsuQC::runVocabChecks()`)
  loops the same way internally -- there's no bulk shortcut even from the inside.
- Some DATRAS-relevant fields (`HaulVal`, `Gear`, `DataType` among them) aren't
  registered under any DATRAS-specific grouping in `icesVocab`'s own
  `findCodeType("DATRAS")` helper, so it wouldn't fully cover this package's fields
  even if adopted.

Separately: some field *descriptions* -- in `getDatrasFieldList()`'s own
`Description` column -- appear to be wrong, occasionally with the hallmarks of a
copy-paste from a different field. Noted directly from working with the data, not
yet catalogued case by case here; worth a dedicated pass at some point, distinct
from the naming/typing gaps above.

## Steps used to find and confirm each gap

The method was the same each time. Code, not just a description of it -- each of
these ran against real DATRAS data during this investigation.

**1. Fetch real data, compare column names against the field list directly.**

```{r}
fl <- getDatrasFieldList()
ca <- getDATRAS("CA", "NS-IBTS", 2022, 1, fix_types = FALSE, new_names = FALSE)

known <- unique(c(fl$FieldNameOld[fl$RecordHeader == "CA"],
                   fl$FieldName[fl$RecordHeader == "CA"]))
setdiff(names(ca), known)
#> [1] "Age"
```

Repeated across HH/HL/CA/FL/LT for the same survey/quarter, `"Age"` in CA was the
*only* column left over with no field-list entry at all -- everything else already
had one.

**2. Repeat with `new_names = FALSE` specifically**, to localise *where* a gap
actually sits.

```{r}
"Age" %in% names(getDATRAS("CA", "NS-IBTS", 2022, 1, new_names = TRUE))
#> [1] TRUE
"Age" %in% names(getDATRAS("CA", "NS-IBTS", 2022, 1, new_names = FALSE))
#> [1] TRUE
```

Same result either way. If the raw, untranslated response already uses a name
absent from the field list's `FieldNameOld` column, the field list's metadata is
what's wrong -- not a translation bug in this package.

**3. Where a real "new headers" endpoint exists, cross-check against it directly**
rather than trusting this package's own translation as ground truth.

```{r}
url_new <- paste0(
  "https://datras.ices.dk/WebServices/DATRASWebService.asmx/getHLdataNewHeaders",
  "?survey=NS-IBTS&year=2022&quarter=1"
)
new_headers   <- parseDatras(readDatras(url_new))
hl_translated <- getDATRAS("HL", "NS-IBTS", 2022, 1, fix_types = FALSE, new_names = TRUE)

setdiff(names(new_headers), names(hl_translated))
#> [1] "Valid_Aphia"    # ICES's real endpoint still uses this name...
setdiff(names(hl_translated), names(new_headers))
#> [1] "aphia"          # ...but this package had invented a different one
```

Without a second, independent source for what ICES actually calls a field, a
locally-invented name can look correct indefinitely.

**4. Treat every fix found this way as interim, not final.** The field list is
ICES's data, not this package's -- a patch here corrects the symptom for this
package's users, but the underlying gap still needs reporting to ICES Datacenter
directly (a different channel to whoever maintains the `icesDatras` R package code
on GitHub -- not yet identified as of writing).
